SU($n$) ferromagnets near the Heisenberg critical point

### 2020-04-10 collective_sun/sun_perturbation.tex
added mention of valence bond solids to SU(n) ferromagnet paper other minor text changes

SU($n$) symmetries play an important role in physics.  Underpinning
much of high energy physics, the celebrated SU($n$) gauge theory known
as Yang-Mills theory is central to our understanding of the
electroweak and strong forces.  Extensions of Yang-Mills and SU($n$)
symmetry feature in the most well-studied examples of holographic
duality\cite{maldacena1999largen} and the connection between
entanglement and gravity\cite{ryu2006holographic} through the anti-de
Sitter / conformal field theory (AdS/CFT) correspondence.  In a
condensed matter setting, SU(2) appears ubiquitously as a symmetry of
the Hubbard model, with important consequences for the study of
quantum magnetism and high temperature
superconductivity\cite{lee2006doping}.  The extension of the SU(2)
Hubbard model to SU($n$) has led to predictions of exotic phases of
matter such as chiral spin liquids\cite{hermele2009mott,
  hermele2011topological, chen2016syntheticgaugefield,
  nataf2016chiral}, as well as the potential to perform universal
topological quantum computation\cite{freedman2004class,
  nayak2008nonabelian}.  Furthermore, the consideration of disordered
SU($n$) spin models has opened analytically tractable avenues for
studying quantum chaos and information
scrambling\cite{sachdev1993gapless, bentsen2019integrable}.

@@

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute all matrix
elements of arbitrary spin operators restricted manifolds of weakly
broken permutation symmetry, thereby enabling the numerical simulation
of dynamics induced by larger, non-perturbative modifications to the
SU($n$) Heisenberg model.  While the latter simulations generally
accumulate uncontrolled truncation errors, we identify criteria under
which numerical results correctly capture the qualitative dynamical
behavior of a system.

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}}
  h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where
$\m s_i\equiv\f12\p{\sigma_\x^{(i)},\sigma_\y^{(i)},\sigma_\z^{(i)}}$
is a spin vector defined with Pauli operators $\sigma_\alpha^{(i)}$
acting on the spin at lattice site $i$.  In the standard language of
spin physics, each term of this Hamiltonian essentially imposes an
energy penalty of $-h_{ij}$ to the formation of a spin 0 singlet state
$\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$, favoring the
formation of spin triplet states with total spin 1.  Another
interpretation of the Hamiltonian in \eqref{eq:H_SU2} is illuminated
by the observation that the two-spin operator $\m s_i\c\m s_j$ is, up
to an overall constant, essentially the two-qubit SWAP operation
$\Pi_{ij}$.  We will also refer to $\Pi_{ij}$ as {\it permutation
  operator}, because its action on the Hilbert space
$\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins with individual Hilbert
spaces $\H_k$ is to permute the tensor factors of $\H$ associated with
spins $i$ and $j$ (see Figure \ref{fig:sketch_spins}).  Up to identity
terms with no physical consequence, we then have
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\substack{\t{spin states}\\\mu,\nu}}
  s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\substack{\t{spin states}\\\mu,\nu}} \op{\mu\nu}{\nu\mu}_{ij},
  \label{eq:H_0}
\end{align}
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$ index two spin states, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index the orthogonal states (sometimes
called {\it colors} or {\it flavors}) of an $n$-level spin.

@@

If the coefficients $h_{ij}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of $H_0$ consists
of spin-polarized, {\it permutationally symmetric} (PS) states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{ij}$.  We will assume that all $h_{ij}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of all spins is a permutationally-symmetric
state.  We will also assume that the PS manifold $\M_0$ is gapped away
from all orthogonal states by an interaction energy
$\Delta_{\t{gap}}>0$.  Power-law couplings of the form
$h_{ij}\sim-1/\abs{\m r_i-\m r_j}^\alpha$ on a $D$-dimensional
lattice, for example, always yield a non-vanishing spectral gap
$\Delta_{\t{gap}}$ for long-range interactions with $\alpha\le D$ (see
Appendix \ref{sec:spectral_gap}).  Here $\m r_i$ denotes the position
of lattice site $i$.

In the case of SU(2), the PS manifold $\M_0$ is precisely the {\it
  Dicke manifold} spanned by the PS states $\ket{m_\up,m_\dn}$ with a
definite (integer) number $m_\mu\ge0$ of spins occupying the
single-spin state $\mu\in\set{\up,\dn}$.  If there are $N$ spins
total, then Dicke states $\ket{m_\up,m_\dn}$ must have
$m_\up+m_\dn=N$.  In the case of multi-level spins, the PS manifold
$\M_0$ is similarly spanned by states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu\ge0$ of the single-spin state $\mu$, with $\sum_\mu m_\mu=N$.
These states are essentially labeled by an assignment
$m=\p{m_1,m_2,\cdots,m_n}$ of $N$ (identical) spins to $n$ (distinct)
states.  We denote the set of all such assignments by $\A_n\p{N}$,
such that our standard basis for $\M_0$ is
$\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$, or
equivalently the ground-state degeneracy of $H_0$, is straightforward
to determine with a standard ``stars and bars'' calculation:
$\dim\M_0=\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.  The
energy of any PS state $\ket\psi\in\M_0$ with respect to the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij}
  = \sum_{i<j} h_{ij},
\end{align}
where $i,j\in\set{1,2,\cdots,N}$ index individual lattice sites.

@@

% give examples of these manifolds in SU(2)

>>>>

SU($n$) symmetries play an important role in physics.  Underpinning
much of high energy physics, the celebrated SU($n$) gauge theory known
as Yang-Mills theory is central to our understanding of the
electroweak and strong forces.  Extensions of Yang-Mills and SU($n$)
symmetry feature in the most well-studied examples of holographic
duality\cite{maldacena1999largen} and the connection between
entanglement and gravity\cite{ryu2006holographic} through the anti-de
Sitter / conformal field theory (AdS/CFT) correspondence.  In a
condensed matter setting, SU(2) appears ubiquitously as a symmetry of
the Hubbard model, with important consequences for the study of
quantum magnetism and high temperature
superconductivity\cite{lee2006doping}.  The extension of the SU(2)
Hubbard model to SU($n$) has led to predictions of exotic phases of
matter such as valence bond solids\cite{read1989valencebond,
  rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and
chiral spin liquids\cite{hermele2009mott, hermele2011topological,
  chen2016syntheticgaugefield, nataf2016chiral}, as well as the
potential to perform universal topological quantum
computation\cite{freedman2004class, nayak2008nonabelian}.
Furthermore, the consideration of disordered SU($n$) spin models has
opened analytically tractable avenues for studying quantum chaos and
information scrambling\cite{sachdev1993gapless,
  bentsen2019integrable}.

@@

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute all matrix
elements of arbitrary spin operators restricted manifolds of weakly
broken permutation symmetry, thereby enabling the numerical simulation
of dynamics induced by non-perturbative modifications to the SU($n$)
Heisenberg model.  While the latter simulations generally accumulate
uncontrolled truncation errors, we identify criteria under which
numerical results correctly capture the qualitative dynamical behavior
of a system.

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}}
  h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where
$\m s_i\equiv\f12\p{\sigma_\x^{(i)},\sigma_\y^{(i)},\sigma_\z^{(i)}}$
is a spin vector defined with Pauli operators $\sigma_\alpha^{(i)}$
acting on the spin at lattice site $i$.  In the standard language of
spin physics, each term of this Hamiltonian essentially imposes an
energy penalty of $-h_{ij}$ to the formation of a spin 0 singlet state
$\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$, favoring the
formation of spin triplet states with total spin 1.  Another
interpretation of the Hamiltonian in \eqref{eq:H_SU2} is illuminated
by the observation that the two-spin operator $\m s_i\c\m s_j$ is, up
to an overall constant, essentially the two-qubit SWAP operation
$\Pi_{ij}$.  We will refer to $\Pi_{ij}$ as {\it permutation
  operator}, because its action on the Hilbert space
$\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins with individual Hilbert
spaces $\H_k$ is to permute the tensor factors of $\H$ associated with
spins $i$ and $j$ (see Figure \ref{fig:sketch_spins}).  Up to identity
terms with no physical consequence, we then have
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\substack{\t{spin states}\\\mu,\nu}}
  s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\substack{\t{spin states}\\\mu,\nu}} \op{\mu\nu}{\nu\mu}_{ij},
  \label{eq:H_0}
\end{align}
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$ index two spin states, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index the orthogonal states (sometimes
called {\it colors} or {\it flavors}) of an $n$-level spin.

@@

If the coefficients $h_{ij}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of $H_0$ consists
of spin-polarized, {\it permutationally symmetric} (PS) states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{ij}$.  We will assume that all $h_{ij}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of all spins is a permutationally-symmetric
state (e.g.~a spin-polarized state).  We will also assume that the PS
manifold $\M_0$ is gapped away from all orthogonal states by an
interaction energy $\Delta_{\t{gap}}>0$.  Power-law couplings of the
form $h_{ij}\sim-1/\abs{\m r_i-\m r_j}^\alpha$ on a $D$-dimensional
lattice, for example, always yield a non-vanishing spectral gap
$\Delta_{\t{gap}}$ for long-range interactions with $\alpha\le D$ (see
Appendix \ref{sec:spectral_gap}).  Here $\m r_i$ denotes the position
of lattice site $i$.

In the case of SU(2), the PS manifold $\M_0$ is precisely the {\it
  Dicke manifold} spanned by the PS states $\ket{m_\up,m_\dn}$ with a
definite (integer) number $m_\mu\ge0$ of spins occupying the
single-spin state $\mu\in\set{\up,\dn}$.  If there are $N$ spins
total, then Dicke states $\ket{m_\up,m_\dn}$ must have
$m_\up+m_\dn=N$.  In the case of multi-level spins, the PS manifold
$\M_0$ is similarly spanned by states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu\ge0$ of the single-spin state $\mu$, with $\sum_\mu m_\mu=N$.
These states are essentially labeled by an assignment
$m=\p{m_1,m_2,\cdots,m_n}$ of $N$ (identical) spins to $n$ (distinct)
states.  We denote the set of all such assignments by $\A_n\p{N}$,
such that our standard basis for $\M_0$ is
$\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$, or
equivalently the ground-state degeneracy of $H_0$, is straightforward
to determine with a standard ``stars and bars'' calculation:
$\dim\p{\M_0}=\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.
The energy of any PS state $\ket\psi\in\M_0$ with respect to the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij}
  = \sum_{i<j} h_{ij},
\end{align}
where $i,j\in\set{1,2,\cdots,N}$ index individual lattice sites.

@@

% todo: give examples of these manifolds in SU(2)
% todo: clarify connection between multi-body excitation manifolds
%       and manifolds of fixed total spin in the SU(2) case
